Changepoint Approach

### 2022-07-26 appendices/changepoint_methods.Rmd
minor edits
finally create pdf version of changepoint methods rmd

---
title: "Changepoint Approach"
author: "Jon Minton"
date: '2022-07-22'
output: html_document
---

@@

Breaks in the data are estimated by fitting the following model specifications:
    * **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
    * **$M_1$**: $\frac{de_x}{dt} \sim \alpha + \beta T$, where

$$
T = \left\{
    \begin{array}\\
        1 & \mbox{if } \ t \ge \tau \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$
    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where

$$
T_1 = \left\{
    \begin{array}\\
        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$

@@

$$
T_2 = \left\{
    \begin{array}\\
        1 & \mbox{if } t \ge \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$
i.e. the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$. As $\tau_1 \lt \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered. Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.

@@

```
# Functions

@@

```{r}
changepoint_breakpoint_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(min_bic = bic == min(bic)) %>% 
  ungroup() %>% 
  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
  geom_point() + 
  facet_wrap(~ sex + x, scales = "free_y") +
  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle')) + 
  labs(
    title = "BIC by proposed changepoint year",
    subtitle = "Red triangle = lowest BIC. Population: England & Wales",
    x = expression(paste("Proposed changepoint year ", tau)),
    y = "BIC (lower is better)",
    caption = "Source: HMD Lifetables"
  ) + 
  theme(legend.position = "none")

@@

```{r}
changepoint_breakpoints_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_taus_from_gridsearch, what = "all")
  )

@@

```
For France, for all four subpopulations, a pairing of `{2004, 2005}` was identified. If we look at the series for France, we can see why this pair of years may have been identified

@@

```
For changes in conditional life expectancy from age 65 two changepoint models with contiguous years were identified for females in Germany, France and Italy; and for males in Scotland, France, Italy and the Netherlands.

@@

The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $tau_2$ conditional on $tau_1$ so as to be two or more years later. (e.g. if $tau_1$ were 1990 then candidate values for $tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.

>>>>

---
title: "Changepoint Approach"
author: "Jon Minton"
date: '2022-07-22'
output:
  pdf_document: default
  word_document: default
  html_document: default
---

@@

Breaks in the data are estimated by fitting the following model specifications:

* **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
* **$M_1$**: $\frac{de_x}{dt} \sim \alpha + \beta T$, where

$$
T = \left\{
    \begin{array}{cc}
        1 & \mbox{if   } \  \ t \ge \tau \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$

* **$M_2$**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where

$$
T_1 = \left\{
    \begin{array}{cc}
        1 & \mbox{if } \tau_1 \le t < \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$

@@

$$
T_2 = \left\{
    \begin{array}{cc}
        1 & \mbox{if } t \ge \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$

i.e. the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t < \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t < \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$. As $\tau_1 < \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered. Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.

@@

# Functions

@@

```{r cache=TRUE}
changepoint_breakpoint_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(min_bic = bic == min(bic)) %>% 
  ungroup() %>% 
  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
  geom_point() + 
  facet_wrap(~ sex + x, scales = "free_y") +
  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle')) + 
  labs(
    title = "BIC by proposed changepoint year",
    subtitle = "Red triangle = lowest BIC. Population: England & Wales",
    x = expression(paste("Proposed changepoint year ", tau)),
    y = "BIC (lower is better)",
    caption = "Source: HMD Lifetables"
  ) + 
  theme(legend.position = "none")

@@

```{r cache=TRUE}
changepoint_breakpoints_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_taus_from_gridsearch, what = "all")
  )

@@

For France, for all four subpopulations, a pairing of `{2004, 2005}` was identified. If we look at the series for France, we can see why this pair of years may have been identified

@@

For changes in conditional life expectancy from age 65 two changepoint models with contiguous years were identified for females in Germany, France and Italy; and for males in Scotland, France, Italy and the Netherlands.

@@

The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $\tau_2$ conditional on $\tau_1$ so as to be two or more years later. (e.g. if $\tau_1$ were 1990 then candidate values for $\tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.
